---
title: "Final Project- Rat sighting by Borough"
author: "Youssra Yemmas"
date: "2023-12-05"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(knitr)
library(janitor)
library(ggplot2)
library(plotly)


knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### One question we are interest in is how do the rat sightings differ over time and across boroughs
```{r}
rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )

summary(rat_sightings)
variable_types <- sapply(rat_sightings, class)
print(variable_types)
# variables are not classified well for analysis so will need too convert numeric variables
numeric_vars_to_convert <- c("unique_key", "month", "year", "incident_zip", "borough_id")

rat_sightings <- rat_sightings %>% 
  mutate(across(all_of(numeric_vars_to_convert), as.factor))
        
variable_types <- sapply(rat_sightings, class)
print(variable_types)

# number of rat sightings by boro each year
rats_boro = rat_sightings %>% 
  janitor::clean_names() %>% 
  select(borough, year, unique_key) %>% 
  group_by(borough, year) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) %>% 
  ungroup %>% 
  spread(key = year, value = avg_rat_sightings) %>% 
  filter(borough != 'Unspecified')# I want to remove the unsepcified

knitr::kable(rats_boro)
  
```

We can see from the kable output that there was quite a substantial jump in rat sightings from 2020 to 2021 in all of the boroughs. This may be another COVID phenomena as restaurants shifted to more outdoor dining which deposited more food waste and other things that attract rats onto the streets during the pandemic and after the pandemic when indoor dining became less feasible. https://apnews.com/article/rats-new-york-9dc65afa66a3535cba01b1ea866973a1#:~:text=NEW%20YORK%20(AP)%20%E2%80%94%20They,so%20did%20the%20city's%20rats.

### Are the differences we can see in average rate sighting across time and boroughs statistically significant? 
```{r}
# I will test the statistical difference of average rat sighting across boroughs
anova1_data =rat_sightings %>% 
  janitor::clean_names() %>% 
  select(borough, unique_key) %>% 
  group_by(borough) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) %>% 
  filter(borough != 'Unspecified')


anova_model1 <- aov(avg_rat_sightings ~ borough, data = anova1_data)

# Print the summary to get F-statistic and p-value
anova1_summary <- summary(anova_model1)
print(anova1_summary)

# Extract F-statistic and p-value
f_statistic <- anova1_summary$`F value`[1]
p_value <- anova1_summary$`Pr(>F)`[1]

# Print the results
cat("F-Statistic:", f_statistic, "\n")
cat("P-Value:", p_value, "\n")

```

```{r}
# I will test the statistical difference of average rat sightings acorss time from 2019-2023

anova2_data = rat_sightings %>% 
  janitor::clean_names() %>% 
  select(year, unique_key) %>% 
  group_by(year) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) 

anova_model2 <- aov(avg_rat_sightings ~ year, data = anova2_data)

# Print the summary to get F-statistic and p-value
anova2_summary <- summary(anova_model2)
print(anova1_summary)

# Extract F-statistic and p-value
f_statistic <- anova2_summary$`F value`[1]
p_value <- anova2_summary$`Pr(>F)`[1]

# Print the results
cat("F-Statistic:", f_statistic, "\n")
cat("P-Value:", p_value, "\n")
```

### Visualizations
```{r}
viz1_data = rats_boro %>% 
  pivot_longer(cols = starts_with("20"),
               names_to = "Year",
               values_to = "avg_rat_sightings"
  )

  ggplot(viz1_data, aes(x = Year, y = avg_rat_sightings)) +
   geom_point(alpha = 0.3, size = 2) +
   geom_line(size = 1, alpha = 0.6) +
   facet_wrap(~borough, scales = "free_y") +
     theme(legend.position = "bottom",
         axis.text.y = element_text(color = "black", 
                                    size = 10,  hjust = 1), 
         axis.text.x = element_text(angle = 45, 
                                    hjust = 1, size = 10)) +
   labs(
     x = "Year",
     y = "Average Rat Sightings",
     title = "Average Rate Sightings From 2019-2023 by Borough"
   ) + 
     viridis::scale_colour_viridis() 

 


```

