---
title: "Exploratory Data Analysis"
output: 
  html_document:
    code_folding: show
---
```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)
library(stats)
library(stats)
library(tidyr)
library(readxl)
library(stringr)

```

```{r import and clean rat sightings, message = FALSE, warning=FALSE}


rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )
```

### Data cleaning and tidying 
In order to analyze rat sightings, we used 2 data sources from NYC Open Data: 311 Rat Sightings and Monthly tonnage data from the NYC Department of Sanitation. The 311 Rat Sightings data was filtered to exclude observations from Staten Island since there was significantly less data from that borough. This decision aims to maintain consistency and reliability in our analysis by focusing on areas with more robust data. Since the data set contained data from more than 230,000 rat sighting reports from 2010 until present-day, we filtered it to only include data from 2019-2023. Only variables of interest to our analysis were kept: 
 - `unique_key`: A unique identifier for each rat sighting report.
 - `location_type`: The type of location where the rat sighting was reported, providing context to the environment of these sightings.
 - `incident_zip`: The ZIP code of the reported sighting, allowing for geographical analysis.
 - `borough`: The borough where the sighting occurred, crucial for comparing sightings across different areas of the city.
 - `location`: Specific coordinates of the sighting, offering detailed geographical information in regards to latitude and longitude.

Variable `created_date` was separated to create three new variables to allow for analysis: `month`, `day`, `year`. 

Borough_id was serves as a numerical identifier for each borough, which can be useful for data management, especially when dealing with categorical data (like the boroughs) in statistical analyses or visualizations. 

```{r data cleaning waste, message = FALSE}
waste_tonnage = read_csv("data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

## Waste Tonnage Data Cleaning

To tidy up the waste tonnage data, we started by importing the Monthly Tonnage Data from the NYC Department of Sanitation. To make the data more manageable and consistent, we first applied the clean_names function. 

Next, we focused on the date information. The month column initially contained combined year and month data, so we split it into two separate variables. We achieved this by using strsplit to separate the year and month, and then applied the sapply function to extract and convert these into individual year and month integer columns. 

We decided to filter the dataset to include only the years 2022 and 2023. This filtering aligns our analysis with a more recent and relevant timeframe and allows for a smaller subset of the initial dataset.

We then consolidated the data on organic waste, which was initially spread across different columns (resorganicstons and schoolorganictons). These represent residential organic tons and school organic tons. We summed these columns to create a new `total_organics` column, giving us a the total organic waste.

For our last step, we grouped the data by borough, month, year, and borough_id. This grouping allowed us to summarize the total organic waste and total refuse collected in each borough for each month and year, ensuring our analysis would capture the geographical and temporal variations in waste tonnage. We used summarise with na.rm = TRUE to handle any missing values effectively, ensuring that they did not skew our results.

```{r merging, include=FALSE}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
rat_waste_filtered = subset(rat_waste_merged, !is.na(total_organics) & !is.na(total_refuse)) #filtering out NAs in the dataset. brianna did this before she did her figures. 
```

Merging Datasets: We used the left_join function from the dplyr package to we merge the rat_sightings data with the waste_tonnage data. We chose to merge based on borough_id, month, year, and borough. This approach ensures that each rat sighting record is aligned with the corresponding waste tonnage data based on both geographical (borough) and temporal (month and year) factors. After merging, we applied a filtering step to remove rows with missing values (NA) in the total_organics and total_refuse columns. 

### Figure 1: Total Trash by Borough for 2022 and 2023
```{r figure 1, message = FALSE}
waste_tonnage_by_borough = rat_waste_filtered |> 
  group_by(borough) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

ggplot(waste_tonnage_by_borough, aes(x = reorder(borough, -total_tonnage), y = total_tonnage)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Borough", y = "Trash Tonnage", title = "Trash Tonnage by Borough, 2022 - 2023") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 1, titled "Trash Tonnage by Borough, 2022 - 2023," is a bar chart visualizing the total amount of trash (comprising both organics and refuse) collected in each borough of New York City over the years 2022 and 2023. From the chart, it is evident that *Brooklyn* has the highest total trash tonnage, shown by the tallest bar. The other boroughs—Queens, Manhattan, and the Bronx—are represented with progressively smaller bars, indicating a decrease in the total trash tonnage respectively. Staten Island is not included in the chart, which is consistent with the earlier note that data from Staten Island was excluded from the analysis due to significantly less data being available from that borough. The scientific notation on the y-axis was used for a concise representation of large numbers, indicating the vast scale of waste collected in across the boroughs. 

### Figure 2: Total Refuse and Total organics 
```{r figure 2, message = FALSE}
waste_by_borough_refuse <- aggregate(total_refuse ~ borough, data=waste_tonnage, sum)

# barplot for total refuse
barplot(waste_by_borough_refuse$total_refuse,
        main="Total Refuse Production by Borough",
        xlab="Borough",
        ylab="Total Refuse (tons)",
        col="blue",
        names.arg=waste_by_borough_refuse$borough,
        las=2)

waste_by_borough_organics <- aggregate(total_organics ~ borough, data=waste_tonnage, sum)

# barplot for total organics
barplot(waste_by_borough_organics$total_organics,
        main="Total Organics Production by Borough",
        xlab="Borough",
        ylab="Total Organics (tons)",
        col="green",
        names.arg=waste_by_borough_organics$borough,
        las=2)
```

- Explain this figure, including why you had to separate (b/c they are on a different scale)

## Figure 3: Trash by Borough per month 
```{r figure 3, message = FALSE}
total_tonnage_by_borough_year_month = rat_waste_filtered |> 
  group_by(borough, year, month) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

total_tonnage_by_borough_year_month$year_month <- as.Date(paste(total_tonnage_by_borough_year_month$year, total_tonnage_by_borough_year_month$month, "01", sep = "-"))

ggplot(total_tonnage_by_borough_year_month, aes(x = year_month, y = total_tonnage, group = borough, color = borough)) +
  geom_line() +
  labs(x = "Year-Month", y = "Amount of Trash (in tons)", title = "Trash Tonnage by Month and Borough") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::number_format(scale = 1e-6)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

- Explain this figure,