---
title: <div align="center"> "Exploratory Data Analysis"
output: 
  html_document:
    code_folding: show
---
```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(dplyr)
library(ggplot2)
library(stats)
library(stats)
library(tidyr)
library(readxl)
library(stringr)
library(patchwork)
library(gridExtra)
library(knitr)
library(tidyverse)
library(janitor)
library(knitr)
library(tidyr)
library(readxl)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(tibble)
library(dbplyr)
library(stats)
library(plotly)
library(ggridges)
library(forecast)
library(patchwork)
library(dplyr)
library(flexdashboard)
library(leaflet)
library(shiny)

library(leaflet.providers)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 7,
  out.width = "90%"
)


theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r import and clean rat sightings, message = FALSE, warning=FALSE}


rat_sightings = 
  read_csv ("./data/Rat_Sightings.csv") |>
  janitor::clean_names(case = "snake") |>
  separate(created_date, sep="/", into = c("month", "day", "year")) |> 
  separate(year, sep=" ", into = c("year")) |>
  filter(borough != "STATEN ISLAND") |> 
  filter(year %in% c("2019", "2020", "2021", "2022", "2023")) |>
  mutate(
    borough_id = recode(
      borough, 
      "MANHATTAN" = 1,
      "BRONX" =2,
      "BROOKLYN"=3,
      "QUEENS"= 4)) |>
  mutate(
    month = as.numeric(month),
    year = as.numeric(year)
  ) |>
  select(unique_key, month, day, year, location_type, incident_zip, borough, location, borough_id) |>
  mutate(
    borough = str_to_sentence(borough)
  )
```

## Data cleaning and tidying 
In order to analyze **the rat sightings** dataset, we used 2 data sources from NYC Open Data: 311 Rat Sightings and Monthly tonnage data from the NYC Department of Sanitation. The 311 Rat Sightings data was filtered to exclude observations from Staten Island since there was significantly less data from that borough. This decision aims to maintain consistency and reliability in our analysis by focusing on areas with more robust data. Since the data set contained data from more than 230,000 rat sighting reports from 2010 until present-day, we filtered it to only include data from 2019-2023. Variable `created_date` was separated to create three new variables to allow for analysis: `month`, `day`, `year`. These new variables and variables of interest to our analysis were kept: 

 * `unique_key`: A unique identifier for each rat sighting report.
 * `location_type`: The type of location where the rat sighting was reported, providing context to the environment of these sightings.
 * `incident_zip`: The ZIP code of the reported sighting, allowing for geographical analysis.
 * `borough`: The borough where the sighting occurred, crucial for comparing sightings across different areas of the city.
 * `location`: Specific coordinates of the sighting, offering detailed geographical information in regards to latitude and longitude.
 * `borough_id`: This variable was created based on `borough` and assigns a numeric value to each of the 4 boroughs represented in our dataset. This was                  done to facilitate the dataset merging later on. 

```{r data cleaning waste, message = FALSE}
waste_tonnage = read_csv("data/DSNY_Monthly_Tonnage_Data_20231202.csv") %>%
  clean_names(case = "snake") %>%
  mutate(date_split = strsplit(month, "/")) %>%
  mutate(
    year = as.integer(sapply(date_split, function(x) x[1])),
    month = as.integer(sapply(date_split, function(x) x[2]))
  ) %>%
  filter(year %in% c(2022, 2023)) %>% 
  mutate(total_organics = resorganicstons + schoolorganictons)

waste_tonnage = waste_tonnage %>% 
  group_by(borough, month, year, borough_id) %>% 
  summarise(
    total_organics = sum(total_organics, na.rm = TRUE),
    total_refuse = sum(refusetonscollected, na.rm = TRUE)
    )
```

The **monthly waste tonnage** dataset was filtered to have observations from 2022 and 2023 so that the effect of the trash pickup timings in early 2023 is evident. In addition, a new variable `total_organics` was created  as the sum of residential organic waste and school organic waste. The dataset was then grouped by `borough`, `month`, `year` and `borough_id` such that each row of the dataset had `total_organics` and `total_refuse` for each month and each borough in 2022 and 2023. Only variables of interest to our analysis were kept: 
  * 
  * 
  * 

```{r merging, include=FALSE}
rat_waste_merged = left_join(rat_sightings, waste_tonnage, by = c("borough_id", "month", "year", "borough"))
rat_waste_filtered = subset(rat_waste_merged, !is.na(total_organics) & !is.na(total_refuse)) #filtering out NAs in the dataset
```

We used the `left_join` function from the `dplyr` package to merge the rat sightings data with the waste tonnage data. We chose to merge based on `borough_id`, `month`, `year`, and `borough`. This approach ensures that each rat sighting record is aligned with the corresponding waste tonnage data based on both geographical (borough) and temporal (month and year) factors. After merging, we applied a filtering step to remove rows with missing values in the `total_organics` and `total_refuse` columns. 

## Visualizations 
### Figure 1: Trash Tonnage by Borough, 2022 - 2023
```{r figure 1, message = FALSE}
waste_tonnage_by_borough = rat_waste_filtered |> 
  group_by(borough) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

ggplot(waste_tonnage_by_borough, aes(x = reorder(borough, -total_tonnage), y = total_tonnage)) +
  geom_bar(stat = "identity", fill = "red3") +
  labs(x = "Borough", y = "Trash Tonnage", title = "Trash Tonnage by Borough, 2022 - 2023") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Figure 1.** This bar chart shows the total amount of trash (comprising both organics and refuse) collected in each borough of New York City in 2022 and 2023. From the chart, it is evident that *Brooklyn* has the highest total trash tonnage, shown by the tallest bar. The other boroughs—Queens, Manhattan, and the Bronx—are represented with progressively smaller bars, indicating a decrease in the total trash tonnage respectively. Staten Island is not included in the chart, as observations from that borough were excluded. The scientific notation on the y-axis was used for a concise representation of large numbers, indicating the vast scale of waste collected in across the boroughs. 

### Figure 2: Total Refuse Production by Boroughs
```{r plot refuse, message = FALSE}

waste_by_borough_refuse <- waste_tonnage %>%
  filter(borough != "Staten Island") %>%
  aggregate(total_refuse ~ borough, data=., sum)

plot_refuse = ggplot(waste_by_borough_refuse, aes(x = reorder(borough, -total_refuse), y = total_refuse)) +
  geom_bar(stat = "identity", fill = "blue4") +
  labs(x = "Borough", y = "Total Refuse (tons)", 
       title = "Total Refuse Production by Borough") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_refuse 

```

**Figure 2.** This plot shows that Brooklyn leads in refuse production, with Queens closely following, indicative of their larger populations and potentially more commercial activity. The Bronx and Manhattan, while still contributing significantly, show lower quantities of refuse. Notably, the volume of refuse is substantially higher than organics, suggesting that non-recyclable and non-compostable waste constitutes a larger proportion of the city’s waste output.

### Figure 3: Total Organics Production by Boroughs
*Refuse and Organic waste are separate on two plots since they have different scales for trash collection.

```{r plot organics, message = FALSE}

waste_by_borough_organics <- waste_tonnage %>%
  filter(borough != "Staten Island") %>%
  aggregate(total_organics ~ borough, data=., sum)

plot_organics = ggplot(waste_by_borough_organics, aes(x = reorder(borough, -total_organics), y = total_organics)) +
  geom_bar(stat = "identity", fill = "green4") +
  labs(x = "Borough", y = "Total Organics (tons)", 
       title = "Total Organics Production by Borough") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


plot_organics 
```

**Figure 3.** This figure portrays a similar bar chart but focuses on organic waste. The trends here are slightly different; Queens surpasses Brooklyn in organic waste, which might reflect more effective organic waste collection policies or higher community engagement in composting programs. Manhattan and the Bronx have much lower figures for organic waste, which might be due to differences in residential patterns, availability of green spaces, or varying levels of participation in organic waste recycling programs.

Comparing the two figures, it is evident that the refuse and organics are managed differently across boroughs, with refuse being the dominant waste type.

### Figure 4: Trash by Borough per month 
```{r figure 3, message = FALSE}
total_tonnage_by_borough_year_month = rat_waste_filtered |> 
  group_by(borough, year, month) |> 
  summarise(total_tonnage = sum(total_organics + total_refuse))

total_tonnage_by_borough_year_month$year_month <- as.Date(paste(total_tonnage_by_borough_year_month$year, total_tonnage_by_borough_year_month$month, "01", sep = "-"))

ggplot(total_tonnage_by_borough_year_month, aes(x = year_month, y = total_tonnage, group = borough, color = borough)) +
  geom_line() +
  labs(x = "Year-Month", y = "Amount of Trash (in tons)", title = "Trash Tonnage by Month and Borough", color = "Borough") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::number_format(scale = 1e-6)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**Figure 4.** This multi-line chart tracks the volume of trash, both organics and refuse, collected over time across different boroughs of New York City from early 2022 to the end of 2023. In this visualization, Brooklyn is depicted with a prominent peak, suggesting a significant increase in trash tonnage in summer months, potentially indicative of seasonal variations or specific events impacting waste production. The other boroughs also show variability, with Queens and Manhattan following similar trajectories and the Bronx displaying the least amount of trash tonnage overall. 

### What are the zip_codes with the highest number of rat sightings?
```{r}
top_zip_codes <- rat_sightings %>%
  group_by(incident_zip) %>%
  summarise(total_sightings = n()) %>%
  arrange(desc(total_sightings)) %>%
  head()
kable(top_zip_codes)

```

The zip codes with the most 311 rat sightings are in the above kable and are associated in order with the areas of the Upper West Side, Bedford- Stuyvesant, Prospect Heights, Bushwick, Flatbush, and Ridgewood. It is interesting that Brooklyn had more zip codes in top six zipcodes of rat sightings but a Manhattan zipcode had the most rat sightings.

### What month had the highest number of rat sightings in each year?

```{r}
# Month with the highest number of rat sightings in each of the five years
top_month_each_year <- rat_sightings %>%
  group_by(year, month) %>%
  summarise(total_sightings = n()) %>%
  arrange(year, desc(total_sightings)) %>%
  group_by(year) %>%
  slice(1)  # Get the top month for each yeartop_month_all_boroughs <- rat_sightings %>%
kable(top_month_each_year)
```

The month with the most rat sightings varied each year so you cannot necesarily see the seasonality of the data just be analyzing the data but when ploting the data by borough the seasonality is much more apparent.

### Switching Focus to Rat Sightings
We were very curious to see how the rat sightings differed across the time period we selected and across boroughs. These sightings are obviously not a measure of actual rat population given that they are based on 311 calls people make but this data is the closest thing we have to a proxy of actual rat populations. Given that, I think a lot of us were biased towards thinking Manhattan would have the most rat sightings and thus by proxy the largest population of rats just based on the density of buildings, people and restaurants. However, following the analysis of trash tonnage it is less suprising that Brooklyn actually leads in rat sightings. 

```{r}
summary(rat_sightings)
variable_types <- sapply(rat_sightings, class)
print(variable_types)
# variables are not classified well for analysis so will need to convert numeric variables
numeric_vars_to_convert <- c("unique_key", "month", "year", "incident_zip", "borough_id")

rat_sightings <- rat_sightings %>% 
  mutate(across(all_of(numeric_vars_to_convert), as.factor))
        
variable_types <- sapply(rat_sightings, class)
print(variable_types)

# number of rat sightings by boro each year
rats_boro = rat_sightings %>% 
  janitor::clean_names() %>% 
  select(borough, year, unique_key) %>% 
  group_by(borough, year) %>% 
  count() %>% 
  summarize(avg_rat_sightings = mean(n)) %>% 
  ungroup %>% 
  spread(key = year, value = avg_rat_sightings) %>% 
  filter(borough != 'Unspecified')# I want to remove the unsepcified
knitr::kable(rats_boro)
```

### Formatting the data in a way that makes plotting easier
```{r}

rat_sightings_agg = rat_sightings |> 
  group_by(year, borough, month) |> 
  filter(borough != "Unspecified") %>% 
  summarise(count = n())
```

We can see from the kable output that there was quite a substantial jump in rat sightings from 2020 to 2021 in all of the boroughs. This may be another COVID phenomena as restaurants shifted to more outdoor dining which deposited more food waste and other things that attract rats onto the streets during the pandemic and after the pandemic when indoor dining became less feasible. https://apnews.com/article/rats-new-york-9dc65afa66a3535cba01b1ea866973a1#. Interestingly, Brooklyn seems to have the highest average of rat sightings amongst the 5 Boros for every year and on the opposite end Queens seems to have the lowest rat sightings pre pandemic but then this year Bronx was the boro with the least rat sightings.


### Figure 5: Rat Sightings Over Time by Borough and Year
```{r}
rat_sightings_agg <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month),
         common_date = lubridate::ym(common_date))

# Plot rat sightings over time
rats_yr_plot <- rat_sightings_agg %>%
  ggplot(aes(x = common_date, y = count, group = year, color = year)) +
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ borough, scales = "free_y", ncol = 2) +  # Facet wrap by borough
  labs(
    title = "Rat Sightings Over Time",
       x = "Date",
       y = "Rat Sightings Count") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "3 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)
ggplotly(rats_yr_plot)
```
**Figure 5.** This plot illustrates the seasonal nature of the rat sightings. we can see an approxsumately normal curve of rat sightings across all boros and years. Across all of the years which are distinguished by the colors the peak of each curve occurs around the spring months of May, June, and July which may have something to do with more people being out and about during those months to encounter the rats. Interestingly, the Bronx experienced its highest number of sightings in 2021 whereas the other three boroughs experienced theirs in 2022.
### Figure 6: Rat Sightings 
```{r}
 bar_plot <- ggplot(rat_sightings_agg, aes(x = year, y = count)) +
  geom_bar(stat = "identity", position = "dodge", fill = "skyblue") +
  facet_wrap(~borough, scales = "free_y") +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(color = "black", size = 10, hjust = 1),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
  ) +
  labs(
    x = "Year",
    y = "Count of Rat Sightings",
    title = "Rat Sightings Across Time and Borough",
  ) +
  viridis::scale_colour_viridis()

ggplotly(bar_plot)
```
**Figure 6.** This bar graph gives a more concrete visual of what was discussed earlier where we can see that although the Bronx generally has less sightings than some of the other boros it experienced an uptick in sightings before the three other boros. 

### Figure 7: Time series of rat sightings
Given the seasonal nature of the rat sightings we saw plotted in the prior plot we wanted to explore how a time series plot would better visualize the data
```{r}
rat_sightings_tibble <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month),
         common_date = lubridate::ym(common_date)) %>%
  as_tibble(index = common_date, key = borough) %>%
  select(borough, common_date, count)  # Include common_date in the select statement

# Time series plot
rats_time_series_plot <- rat_sightings_tibble %>%
  ggplot(aes(x = common_date, y = count, color = borough, group = borough)) +
  geom_line() +
  labs(title = "Rat Sightings Time Series",
       x = "Date",
       y = "Rat Sightings Count",
       color = "Borough") +
  theme_minimal()

# Print the plot
ggplotly(rats_time_series_plot)
```

**Figure 7.** This figure further cements the seasonal nature of the rat sightings and does a better job at depicting on the same scale how much the rat sightings jumped following the pandemic. We can also see from this plot that the trash time rule that occured on 04/01/2023 did not make a noticeable impact on the rat sightings and we are assuming by proxy on the rat population.

### Figure 7: Exploring the forecasting of rat sightings in Brooklyn and the impact of the April 1st Trash Time Change
Brooklyn has the highest average of rat sightings so a time series decomposition plot with forcasting may shed some light into how the rat sightings would have looked if the trash time rule was not implemented. 
```{r}
borough_data <- rat_sightings_tibble %>%
  filter(borough == "Brooklyn")  

# Convert data to a univariate time series
ts_data <- ts(borough_data$count, frequency = 12)  
# Time series analysis using STL decomposition
ts_analysis <- stlf(ts_data, h = 12)  

# Plot time series decomposition
autoplot(ts_analysis) +
  labs(title = "Time Series Decomposition",
       y = "Rat Sightings Count",
       color = "Components") +
  theme_minimal()

rat_sightings_tibble2 <- rat_sightings_agg %>%
  mutate(common_date = paste0(year, "-", month, "-01"),
         common_date = lubridate::ymd(common_date),
         rule_impact = ifelse(common_date >= "2023-04-01", "After", "Before")) %>%
  as_tibble(index = common_date, key = c(borough, rule_impact)) %>%
  select(borough, common_date, count, rule_impact)

rats_rule_impact_plot <- rat_sightings_tibble2 %>%
  ggplot(aes(x = common_date, y = count, color = rule_impact, group = interaction(borough, rule_impact))) +
  geom_line() +
  labs(title = "Impact of Rat Extermination Rule",
       x = "Date",
       y = "Rat Sightings Count",
       color = "Rule Impact",
       subtitle = "Comparison Before and After April 2013") +
  theme(text = element_text(size = 15), 
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10)) +
  scale_colour_discrete("Year") +
  scale_x_date(date_breaks = "3 month", labels = function(x) format(x, "%b")) +
  scale_color_viridis_d(end = .8)

print(rats_rule_impact_plot)
```